Sperm should evolve to make female meiosis fair.

### 2014-05-20 writeup_and_notes/male_dependent_drive.tex
fixed minor tyos in appendix

\subsection*{Models 1-3. Single-locus drive}
\subsubsection*{Model 1. Traditional driver}
In the standard female drive model, meiosis in males is fair such that $A/B$ heterozygotes contribute $A$ and $B$ alleles with equal probabilities; however, $A/B$ females transmit the $B$ allele with probability $d>1/2$.  
We note that the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}. 
After drive and random mating, genotype frequencies are 
	$f_{AA}'=f_A (f_{AA }+ f_{AB} (1 - d))$, 
	$f_{AB}'=f_A (f_{BB} + f_{ABd}) + f_B (f_{AA} + f_{AB} (1 - d))$,
	and $f_{BB}= f_B (f_{AB} d + f_{BB})$. 
%Assuming fitnesses $w_{AA}=1$, $w_{AB}-1 s_h$ and $w_{BB}=1-s$, 
%Frequencies of genotype, $i$, after drive and selection are 
%	\begin{equation} f_i''= f_i'w_i/\bar{w}\end{equation}
%	where $w_i$ is genotypic fitness, and mean fitness, $\bar{w} = \Sigma f_i'w_i$. 
As detailed above, exact frequencies after drive, random mating and selection are $f_g''= f_g'w_g/\bar{w}$. 
Assuming HWE, a rare driver will spread when $s_h
        \lessapprox(2d-1)/(1+2d)$, and will fix when $s\lessapprox d -1/2 +
        3 s_h /2- d s_h$. This later inequality reduces to
        $s\lessapprox(2d-1)/2$ when the cost of drive is fully
        recessive. 
        %\gc{isn't this last expression exact?} YB: exact-ish.. all assume HWE.. but i think we don't expect a deviation from HWE here??  it's prob easiest to kept them all as approx?

\subsubsection*{Model 2. Single locus, sperm-dependent drive.}
Our single-locus model of sperm-dependent drive resembles the traditional driver, with the caveat that the $B$ allele drives in heterozygous females only when fertilized by $B$-bearing sperm. 
Therefore, genotype frequencies after drive are 
	$f_{AA}'=f_A^2$, 
	$f_{AB}'=f_A f_B+f_B (f_{AA} + f_{AB}(1 - d)) $,
	and $f_{BB}'= f_B (f_{AB} d + f_{BB} )$. 
%As above, exact frequencies after drive, random mating and selection are $f_i''= f_i'w_i/\bar{w}$. 
We iterate exact genotype frequency recursions ($f_g''= f_g'w_g/\bar{w}$) over generations to produce the frequency trajectories shown in the inset
Figure \ref{IEggsperm}B by plotting $f_B=f_{BB}+ \frac{1}{2}f_{AB}$
over time. %, and to assess the analytical approximations, below.
To assess invasion or fixation criteria, as well as bistability points, we iterate this system and
test whether $f_B$ increases over a grid of parameters.

@@

\subsubsection*{Model 3. Single locus, paternal genotype dependent drive:}
In the case when female meiotic drive depends on paternal genotype
%\paragraph{Dominance} we also explored a model where
%the genotype of the male who produced the fertilizing sperm affects female drive,
%instead of just the allelic type of the sperm.
 %We assume a 
 heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $d_{het}\geq 1/2 $ or $d_{hom}\geq d_{het}$, 
%YB RE-PARAMETERIZING  %  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$,  YB RE-PARAMETERIZING 
 when mated with to $AA$, $AB$, or $BB$ males,  respectively. 
  In this model, genotype frequencies after drive, random mating, and selection are 
  	$f_{AA}'' =   w_{AA}\left( f_{AA} [2 f_A + f_{AB} ] + f_{AB}^2 [1 - d_{het}] \right)/(2\bar{w})$,
	$f_{AB}'' =   w_{AB}(f_{AA} [f_B + f_{AB}/2] + f_{AB} f_B + f_{BB} [ f_A - f_{AB} d_{hom}])/\bar{w}$, 
	and 	$f_{AB}'' =   w_{AB}(f_{AB} (f_{AB} d_{het}/2 + f_{BB} d_{hom}) + f_{BB} f_B)/\bar{w}$. 
% Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
%We set up a set of recursions using a recessive sex independent cost to
%the driver, similar to those in eqn.
%\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
%transmission probabilities.

>>>>

\subsection*{Models 1-3. Single-locus drive}
\subsubsection*{Model 1. Traditional driver}
In the standard female drive model, meiosis in males is fair such that $A/B$ heterozygotes contribute $A$ and $B$ alleles with equal probabilities; however, $A/B$ females transmit the $B$ allele with probability $d>1/2$.  
We note that the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}. 
After drive and random mating, genotype frequencies are 
	$f_{AA}'=f_A (f_{AA }+ f_{AB} (1 - d))$, 
	$f_{AB}'=f_A (f_{BB} + f_{AB}d) + f_B (f_{AA} + f_{AB} (1 - d))$,
	and $f_{BB}'= f_B (f_{AB} d + f_{BB})$. 
%Assuming fitnesses $w_{AA}=1$, $w_{AB}-1 s_h$ and $w_{BB}=1-s$, 
%Frequencies of genotype, $i$, after drive and selection are 
%	\begin{equation} f_i''= f_i'w_i/\bar{w}\end{equation}
%	where $w_i$ is genotypic fitness, and mean fitness, $\bar{w} = \Sigma f_i'w_i$. 
As detailed above, exact frequencies after drive, random mating and selection are $f_g''= f_g'w_g/\bar{w}$. 
Assuming HWE, a rare driver will spread when $s_h
        \lessapprox(2d-1)/(1+2d)$, and will fix when $s\lessapprox d -1/2 +
        3 s_h /2- d s_h$. This later inequality reduces to
        $s\lessapprox(2d-1)/2$ when the cost of drive is fully
        recessive. 
        %\gc{isn't this last expression exact?} YB: exact-ish.. all assume HWE.. but i think we don't expect a deviation from HWE here??  it's prob easiest to kept them all as approx?

\subsubsection*{Model 2. Single locus, sperm-dependent drive.}
Our single-locus model of sperm-dependent drive resembles the traditional driver, with the caveat that the $B$ allele drives in heterozygous females only when fertilized by $B$-bearing sperm. 
Therefore, genotype frequencies after drive are 
	$f_{AA}'=f_A^2$, 
	$f_{AB}'=f_A f_B+f_B (f_{AA} + f_{AB}(1 - d)) $,
	and $f_{BB}'= f_B (f_{AB} d + f_{BB} )$. 
%As above, exact frequencies after drive, random mating and selection are $f_i''= f_i'w_i/\bar{w}$. 
We iterate exact genotype frequency recursions ($f_g''= f_g'w_g/\bar{w}$) over generations to produce the frequency trajectories shown in the inset of 
Figure \ref{Eggsperm}B by plotting $f_B=f_{BB}+ \frac{1}{2}f_{AB}$
over time. %, and to assess the analytical approximations, below.
To assess invasion or fixation criteria, as well as bistability points, we iterate this system and
test whether $f_B$ increases over a grid of parameters.

@@

\subsubsection*{Model 3. Single locus, paternal genotype dependent drive:}
In the case when female meiotic drive depends on paternal genotype
%\paragraph{Dominance} we also explored a model where
%the genotype of the male who produced the fertilizing sperm affects female drive,
%instead of just the allelic type of the sperm.
 %We assume a 
 heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $d_{het}\geq 1/2 $ or $d_{hom}\geq d_{het}$, 
%YB RE-PARAMETERIZING  %  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$,  YB RE-PARAMETERIZING 
 when mated with to $AA$, $AB$, or $BB$ males,  respectively. 
  In this model, genotype frequencies after drive, random mating, and selection are 
  	$f_{AA}'' =   w_{AA}\left( f_{AA} [2 f_A + f_{AB} ] + f_{AB}^2 [1 - d_{het}] \right)/(2\bar{w})$,
	$f_{AB}'' =   w_{AB}(f_{AA} [f_B + f_{AB}/2] + f_{AB} f_B + f_{BB} [ f_A - f_{AB} d_{hom}])/\bar{w}$, 
	and 	$f_{BB}'' =   w_{BB}(f_{AB} (f_{AB} d_{het}/2 + f_{BB} d_{hom}) + f_{BB} f_B)/\bar{w}$. 
% Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
%We set up a set of recursions using a recessive sex independent cost to
%the driver, similar to those in eqn.
%\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
%transmission probabilities.
